---
title: "S3_Exploratory_Data_Analysis"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exploratory Data Analysis

Brief description:
Exploratory data analysis (EDA) is an approach to analyzing data sets to summarize their main characteristics, 
often with visual methods. A statistical model can be used or not, but primarily EDA is for seeing what the 
data can tell us beyond the formal modeling or hypothesis testing task. 


## 0. Load the libraries

```{r Load the libraries, echo=TRUE}

# Load the ggplot2 library
library(ggplot2)
library(ggpubr) 

```


## 1. Load the data

```{r Load the data, echo=TRUE}

# Load the CHD gene annotation data
chd_genes <- read.delim("../../data/chd_genes.annotations.tsv", header = TRUE, sep = "\t")

# print the head of the data frame
head(chd_genes)

```

\newpage

## 2. Summarize the data

```{r Summarize the data, echo=TRUE}

# Summarize the data

# Mean of the pLI score
mean(chd_genes$pLI) 

# Median of the pLI score
median(chd_genes$pLI)

# Standard deviation of the pLI score
sd(chd_genes$pLI)

# Maximum of the pLI score
max(chd_genes$pLI)

# Minimum of the pLI score
min(chd_genes$pLI)

# Summary of the pLI score
summary(chd_genes$pLI)

```

\newpage

## 3. Visulization of the data using ggplot2

### 3.1 Histogram of gene length

*Histogram*: A histogram is an accurate representation of the distribution of numerical data.

```{r Histogram of gene length, echo=TRUE}

# Histogram of gene length
ggplot(chd_genes, aes(x = gene_length)) + 
  geom_histogram() +
  labs(title = "Histogram of gene length", x = "Gene length", y = "Count")

```


### 3.2 Boxplot of plof per category

*Boxplot*: A boxplot is a standardized way of displaying the distribution of data based on a five number summary:
(minimum, first quartile (Q1), median, third quartile (Q3), and maximum).

```{r Boxplot of plof per category, echo=TRUE}

# Boxplot per category
ggplot(chd_genes, aes(x = category, y = plof, fill=category)) + 
  geom_boxplot() +
  labs(title = "Boxplot of plof per category", x = "category", y = "plof")


```


\newpage

### 3.3 Scatter plot of obs_syn vs obs_lof

*Scatter plot*: A scatter plot (also called a scatterplot, scatter graph, scatter chart, scattergram, or scatter diagram)
is a type of plot or mathematical diagram using Cartesian coordinates to display values for typically two variables.

```{r Scatter plot of obs_syn vs obs_lof, echo=TRUE}

# Scatter plot
ggplot(chd_genes, aes(x = obs_syn, y = obs_lof)) + 
  geom_point() +
  labs(title = "Scatter plot of obs_syn vs obs_lof")

```

### 3.4 Scatter plot with regression line and correlation coefficient

```{r Scatter plot with regression line and correlation coefficient, echo=TRUE}

# Scatter plot with regression line and correlation coefficient
ggplot(chd_genes, aes(x = obs_syn, y = exp_syn)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Scatter plot with regression line") +
  stat_cor(method = "pearson", label.x = 100, label.y = 1500)

```


\newpage

### 3.5 Density plot per category for plof metric

*Density plot*: A density plot is a smoothed, continuous version of a histogram estimated from the data.

```{r Density plot per category for plof metric, echo=TRUE}

# Density plot per category for plof metric
ggplot(chd_genes, aes(x = plof, fill = category)) + 
  geom_density(alpha = 0.5) +
  labs(title = "Density plot")


```


### 3.6 Bar plot

```{r Bar plot - count number of genes per category, echo=TRUE}

# count number of genes per category
ggplot(chd_genes, aes(x = category, fill = category)) + 
  geom_bar() +
  labs(title = "Bar plot of number of genes per category")

# count number of genes per chromosome
ggplot(chd_genes, aes(x = chromosome)) + 
  geom_bar() +
  labs(title = "Bar plot of number of genes per chromosome")

```


\newpage

## 4. Statistical tests

*T-test*: A t-test is a type of inferential statistic used to determine if there is a significant difference between
the means of two groups, which may be related in certain features.

*Wilcoxon test*: The Wilcoxon test, which refers to either the Rank Sum test or the Signed Rank test, is a non-parametric
test that compares two groups. It is the non-parametric equivalent of the t-test.

```{r Statistical tests, echo=TRUE}


# t.test
t.test(pLI ~ category, data = chd_genes, alternative = "two.sided")

# wilcox.test
wilcox.test(pLI ~ category, data = chd_genes, alternative = "two.sided")

```

\newpage

## 5. Boxplot with statistical test

Note: A R-package can be extended to add new functionalities. The ggpubr package extends ggplot2 functionalities.

```{r Boxplot with statistical test, echo=TRUE}

ggboxplot(chd_genes, 
          x = "category", 
          y = "plof",
          color = "category", 
          palette = "jco") +
  stat_compare_means(method = "wilcox.test", label.y = 2) +
  labs(title = "Boxplot of plof by category")

```

## 6. R session information

```{r R session information, echo=TRUE}

sessionInfo()

```