---
title: "S7_Orthologs_Networks_Pathways"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

S6 ended on a failure. Three hundred and fifty-four genes go up, strongly and
significantly, and they enrich in nothing at all.

This lesson asks whether those genes are really uninformative, or whether we
were simply asking a database that does not know very much.


## 1. Load libraries

```{r load libraries, echo = TRUE, message=FALSE, warning=FALSE}

library(dplyr, quietly = TRUE)
library(readr, quietly = TRUE)
library(ggplot2, quietly = TRUE)

```

```{r rebuild lists, echo = TRUE}

# The same lists S6 built.
res = read_tsv("../../results/DESeq2_results.tsv", show_col_types = FALSE)

de = res %>% filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1)

genes_up   = sort(de$gene[de$log2FoldChange > 0])
genes_down = sort(de$gene[de$log2FoldChange < 0])

```


## 2. How much does the database actually know?

FishEnrichr's GO Biological Process library is dated **2018**. The human one is
current. That is not an accident of this website: zebrafish annotation really is
thinner and older than human annotation, because far more people have spent far
more money studying human genes.

So an alternative explanation for S6's empty result is available. Perhaps the
up-regulated genes do something specific, and nobody has written it down in
zebrafish.

We can test that. Most zebrafish genes have a human counterpart.


## 3. Orthologs

Two genes are **orthologs** if they descend from the same gene in the common
ancestor of the two species. Ensembl computes these for every pair of species it
hosts, and labels each relationship:

| Type | Meaning |
| --- | --- |
| `ortholog_one2one` | one zebrafish gene, one human gene |
| `ortholog_one2many` | the relationship is one-to-many in some direction |
| `ortholog_many2many` | several genes on both sides |

Zebrafish is an awkward case. Its lineage went through an extra whole-genome
duplication, so a single human gene frequently has two zebrafish copies, named
`a` and `b`: human `COL1A1` corresponds to zebrafish `col1a1a` and `col1a1b`.

### From R

```{r biomart, echo = TRUE, eval = FALSE}

library(biomaRt)

mart = useEnsembl("genes", "drerio_gene_ensembl")

orth = getBM(
  attributes = c("external_gene_name",
                 "hsapiens_homolog_associated_gene_name",
                 "hsapiens_homolog_orthology_type",
                 "hsapiens_homolog_orthology_confidence"),
  mart = mart)

head(orth)

```

This chunk is not run when the lesson is knitted, and you may well find it fails
when you run it. `biomaRt` needs a live connection to Ensembl and up-to-date
certificates on your machine; while this course was being written it failed with
`SSL certificate problem: certificate has expired`.

That is ordinary. It is also why `data/enrichment/make_enrichment_data.R`
carries a fallback that asks the same Ensembl for the same four columns over
plain HTTP, and why the table is committed to the repository.

### From the browser

The same query, by hand: <https://www.ensembl.org/biomart/martview>

- **Dataset**: Ensembl Genes → Zebrafish genes
- **Attributes**: uncheck the defaults, then under *Homologues* pick
  Gene name, Human gene name, Human homology type
- **Results** → Export all results to TSV

### The committed table

```{r load orthologs, echo = TRUE}

orth = read_tsv("../../data/enrichment/zfish_human_orthologs.tsv",
                show_col_types = FALSE)

head(orth)

nrow(orth)

n_distinct(orth$zfish_symbol)

```


## 4. What "1:1" has to mean

The obvious filter is Ensembl's own label. It is also wrong here, and the reason
is worth the five minutes.

```{r the trap, echo = TRUE}

orth %>%
  filter(zfish_symbol %in% c("serpinh1b", "col1a1a", "col1a1b", "hspa5", "col1a2")) %>%
  select(zfish_symbol, human_symbol, orthology_type, n_human_partners)

```

`serpinh1b` is labelled `ortholog_one2many`, so a strict `ortholog_one2one`
filter throws it away. But look at `n_human_partners`: it is **1**. `serpinh1b`
maps to exactly one human gene, `SERPINH1`. The label says "many" because
*`SERPINH1` has two zebrafish partners*, `serpinh1a` and `serpinh1b`.

The ambiguity is on the zebrafish side. We are travelling from zebrafish to
human, so it does not affect us.

This matters because `serpinh1b` is the most strongly induced gene in the whole
experiment:

```{r serpinh1b, echo = TRUE}

res %>% filter(gene == "serpinh1b") %>% select(gene, baseMean, log2FoldChange, padj)

```

A strict `ortholog_one2one` filter would silently delete it, along with
`col1a1a`, `col1a1b`, `col2a1a`, `ankrd1b` and every other gene the teleost
duplication touched — which, in a zebrafish experiment, is a great many of the
interesting ones.

So the filter we want is "one zebrafish gene, one human gene", expressed
directly:

```{r unambiguous, echo = TRUE}

unambiguous = orth %>% filter(n_human_partners == 1)

to_human = function(fish_genes) {
  sort(unique(unambiguous$human_symbol[unambiguous$zfish_symbol %in% fish_genes]))
}

human_up   = to_human(genes_up)
human_down = to_human(genes_down)

length(genes_up);   length(human_up)
length(genes_down); length(human_down)

```

Several zebrafish genes may still land on the same human gene — `col1a1a` and
`col1a1b` both give `COL1A1`. That direction is safe: we take unique symbols, so
a gene set can only shrink, never inflate.

The next section shows what happens when it is allowed to inflate.


## 5. What the mapping finds

```{r write human lists, echo = TRUE}

dir.create("outputs", showWarnings = FALSE)

writeLines(human_up,   "outputs/genes_up_human.txt")
writeLines(human_down, "outputs/genes_down_human.txt")

```

Submit these to **Enrichr** (the human site, <https://maayanlab.cloud/Enrichr/>)
against **GO Biological Process 2025**, exactly as in S6. Or read the committed
results:

```{r read human enrichment, echo = TRUE}

read_enrichment = function(mine, committed) {
  path = if (file.exists(mine)) mine else committed
  read_tsv(path, show_col_types = FALSE)
}

human_up_go = read_enrichment("outputs/GO_Biological_Process_2025_table.txt",
                              "../../data/enrichment/human_GO_BP_2025_up.tsv")

human_up_go %>%
  filter(`Adjusted P-value` < 0.05) %>%
  select(Term, Overlap, `Adjusted P-value`, Genes)

```

In zebrafish these genes enriched in nothing. In human they carry the
**unfolded protein response**: `HSPA5` is BiP, the master sensor of endoplasmic
reticulum stress, and `PDIA4`, `PDIA6`, `HSP90B1` and `HYOU1` are all ER folding
machinery.

And this came from *fewer* genes — 131 human symbols out of 354 zebrafish ones.
The limitation was never the statistics. It was how much had been written down.

### The control

A result that appears only after a transformation deserves suspicion. The
transformation might have created it.

The test is the down-regulated list, which already had a clear answer in
zebrafish. If the mapping is trustworthy, it should give the same answer again.

```{r human down, echo = TRUE}

human_down_go = read_enrichment("outputs/GO_Biological_Process_2025_table_down.txt",
                                "../../data/enrichment/human_GO_BP_2025_down.tsv")

human_down_go %>%
  filter(`Adjusted P-value` < 0.05) %>%
  select(Term, Overlap, `Adjusted P-value`) %>%
  head(8)

```

Extracellular matrix organisation and collagen fibril organisation, as before.
The mapping reproduces what we already knew, so the new finding in the up-list
is more believable.

Notice also what the richer annotation adds: `Metal Ion Transport` and
`Modulation of Chemical Synaptic Transmission` were invisible in the 2018
zebrafish library.


## 6. What happens without the filter

Suppose we had kept every ortholog, ambiguous ones included.

```{r allortho, echo = TRUE}

all_up_go = read_tsv("../../data/enrichment/human_GO_BP_2025_up_ALLORTHO.tsv",
                     show_col_types = FALSE)

all_up_go %>%
  filter(`Adjusted P-value` < 0.05) %>%
  select(Term, Overlap, `Adjusted P-value`) %>%
  head(5)

```

`Cellular Response to Zinc Ion` at an adjusted p-value of 2 in 10^19. Sixty
significant terms instead of two. A spectacular result.

It is also entirely fabricated. Look at the genes:

```{r allortho genes, echo = TRUE}

all_up_go$Genes[1]

```

```{r mt2, echo = TRUE}

orth %>% filter(zfish_symbol == "mt2") %>% select(zfish_symbol, human_symbol, orthology_type)

```

Twelve of those thirteen genes are human metallothioneins, and every one of them
entered the list through a **single** zebrafish gene, `mt2`. One differentially
expressed gene was multiplied into twelve, and the hypergeometric test — which
assumes the genes are independent draws — believed it.

The same happens on the down side, where four `ugt` genes expand to nineteen
human UGTs each and manufacture a steroid metabolism story.

> **An ortholog mapping that expands genes will invent findings. Check how many
> partners each gene brought with it.**


## 7. The same list, four GO releases

We have been reading "GO Biological Process 2025" as though GO were a fixed
object. It is not. Here is the identical 131-gene list against four releases.

```{r vintages, echo = TRUE}

vintages = c(2021, 2023, 2025, 2026)

summary_table = lapply(vintages, function(year) {
  tb = read_tsv(sprintf("../../data/enrichment/human_GO_BP_%d_up.tsv", year),
                show_col_types = FALSE)
  sig = tb %>% filter(`Adjusted P-value` < 0.05) %>% arrange(`Adjusted P-value`)
  data.frame(
    release     = year,
    terms_tested = nrow(tb),
    significant  = nrow(sig),
    top_term     = sub(" \\(GO:.*", "", sig$Term[1]),
    top_padj     = signif(sig$`Adjusted P-value`[1], 2)
  )
})

do.call(rbind, summary_table)

```

Eighteen significant terms in 2021, two in 2025, eleven in 2026 — on identical
input, from identical code. Had you run this analysis in 2021 you would have had
a much richer story to tell, and most of it would since have evaporated.

One thing does not move:

```{r upr across vintages, echo = TRUE}

upr = lapply(vintages, function(year) {
  tb = read_tsv(sprintf("../../data/enrichment/human_GO_BP_%d_up.tsv", year),
                show_col_types = FALSE)
  row = tb %>% filter(grepl("unfolded protein", Term, ignore.case = TRUE)) %>%
        arrange(`Adjusted P-value`) %>% head(1)
  data.frame(release = year,
             overlap = row$Overlap,
             padj    = signif(row$`Adjusted P-value`, 2))
})

do.call(rbind, upr)

```

The unfolded protein response is significant in every release. Even the size of
the GO term itself changes between versions, and the result survives anyway.

That is the lesson, and it generalises well beyond GO:

> **A conclusion that survives the database version is robust. One that does not
> is a property of the annotation, not of your experiment.**

Re-running against another release costs one changed string. It is the cheapest
robustness check available.


## 8. Networks: STRING

Enrichment asks whether a list is full of a known category. A network asks a
different question: do these proteins touch each other?

Go to <https://string-db.org>, choose **Multiple proteins**, paste the contents
of `outputs/genes_down_human.txt`, and set the organism to *Homo sapiens*.

The edges are also available as a table:

```{r string, echo = TRUE}

edges = read_tsv("../../data/enrichment/string_edges_human_down.tsv",
                 show_col_types = FALSE)

nrow(edges)

edges %>% select(preferredName_A, preferredName_B, score) %>% head()

```

```{r string hubs, echo = TRUE}

# Which proteins have the most connections?
hubs = c(edges$preferredName_A, edges$preferredName_B) %>%
  table() %>%
  sort(decreasing = TRUE) %>%
  head(12)

hubs

```

**Question.** Compare the hub list with the genes behind the top GO term in
section 5. Do the two methods agree? Should they?


## 9. Pathways: Reactome

Reactome is a curated map of human pathways, more detailed and more strictly
maintained than GO.

Paste `outputs/genes_down_human.txt` into <https://reactome.org> → *Analyse gene
list*, or read the committed result:

```{r reactome, echo = TRUE}

reactome = read_tsv("../../data/enrichment/human_Reactome_2024_down.tsv",
                    show_col_types = FALSE)

reactome %>%
  filter(`Adjusted P-value` < 0.05) %>%
  select(Term, Overlap, `Adjusted P-value`) %>%
  head(10)

```

**Question.** Reactome and GO are built by different curators from the same
literature. Where do they agree here, and where does one of them see something
the other does not?


## 10. Exercise: put the story together

You now have, for the same experiment:

- collagen and extracellular matrix genes strongly **down**
- endoplasmic reticulum and protein-folding machinery strongly **up**
- `serpinh1b` / `SERPINH1` up more than thirteen-fold

```{r serpinh1 identity, echo = TRUE}

res %>%
  filter(gene %in% c("serpinh1b", "hspa5", "hsp90b1", "pdia4",
                     "col1a1a", "col1a2", "col11a2")) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(log2FoldChange)

```

**Question.** `SERPINH1` is better known as **HSP47**, and it is not a general
chaperone: it folds collagen, and nothing else.

Write two sentences connecting the three observations above into one
explanation. Then write a third sentence describing an experiment that would
test it.

**Second question.** Everything in this lesson rests on 131 of 354
up-regulated genes. What happened to the other 223, and could their loss have
biased the answer?


## 11. Print R session info

```{r print out R session info, echo = TRUE}
# Always good practice to print out the R session info for reproducibility
sessionInfo()
```
