---
title: "S7_Orthologs_Networks_Pathways"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

In S6, the 354 up-regulated genes showed no significant enrichment in the
zebrafish GO library. This lesson maps the zebrafish genes to human orthologs,
repeats the enrichment with human annotation, and introduces network (STRING)
and pathway (Reactome) analysis.


## 1. Load libraries

```{r load libraries, echo = TRUE, message=FALSE, warning=FALSE}

library(dplyr, quietly = TRUE)
library(readr, quietly = TRUE)
library(ggplot2, quietly = TRUE)

```

```{r rebuild lists, echo = TRUE}

# The same lists built in S6.
res = read_tsv("../../results/DESeq2_results.tsv", show_col_types = FALSE)

de = res %>% filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1)

genes_up   = sort(de$gene[de$log2FoldChange > 0])
genes_down = sort(de$gene[de$log2FoldChange < 0])

```


## 2. Annotation coverage

The GO Biological Process library in FishEnrichr dates from **2018**, while the
human libraries in Enrichr are updated regularly. Zebrafish genes are also, in
general, less extensively annotated than their human counterparts.

A lack of enrichment can therefore have two explanations: the genes do not share
a common function, or their function is not yet recorded in the zebrafish
annotation. The second possibility can be tested by mapping the genes to human
orthologs and repeating the analysis.


## 3. Orthologs

Two genes are **orthologs** if they descend from a single gene in the last common
ancestor of the two species. Ensembl provides ortholog predictions between
species and classifies each relationship:

| Type | Meaning |
| --- | --- |
| `ortholog_one2one` | one zebrafish gene, one human gene |
| `ortholog_one2many` | one gene in one species, several in the other |
| `ortholog_many2many` | several genes in both species |

The teleost lineage underwent an additional whole-genome duplication, so a
single human gene often has two zebrafish copies, usually suffixed `a` and `b`.
For example, human `COL1A1` corresponds to zebrafish `col1a1a` and `col1a1b`.

### Using R

```{r biomart, echo = TRUE, eval = FALSE}

library(biomaRt)

mart = useEnsembl("genes", "drerio_gene_ensembl")

orth = getBM(
  attributes = c("external_gene_name",
                 "hsapiens_homolog_associated_gene_name",
                 "hsapiens_homolog_orthology_type",
                 "hsapiens_homolog_orthology_confidence"),
  mart = mart)

head(orth)

```

This chunk is not evaluated when the lesson is knitted, and it may fail if you
run it. `biomaRt` needs Ensembl's service to be reachable and to be where the
installed version of the package expects it, and that is not always the case.
Two errors seen while preparing this course:

- `SSL certificate problem: certificate has expired`
- `Error in bmRequest(...) : Not Found (HTTP 404)`

Neither means you did anything wrong. Ensembl changed where this service lives,
and older versions of `biomaRt` look in the old place. It is worth seeing once,
because depending on an outside service is normal in this kind of analysis, and
so is having it fail.

There are two ways on, and the lesson uses both:

1. Run the query on the Ensembl website instead, as described next.
2. Use the ortholog table stored in `data/enrichment/`, which is what the rest
   of this lesson reads. It was produced by the same query.

### Using the website

The same query can be run at <https://www.ensembl.org/biomart/martview>:

- **Dataset**: Ensembl Genes → Zebrafish genes
- **Attributes**: clear the defaults, then under *Homologues* select
  *Gene name*, *Human gene name* and *Human homology type*
- **Results** → export all results to TSV

The stored table additionally contains `n_human_partners`, computed by the
generator script, which is used below.

### Load the stored table

```{r load orthologs, echo = TRUE}

# import the stored ortholog table, which was generated by data/enrichment/make_enrichment_data.R
orth = read_tsv("../../data/enrichment/zfish_human_orthologs.tsv",
                show_col_types = FALSE)

head(orth)

nrow(orth)

# print the number of unique zebrafish genes in the table
n_distinct(orth$zfish_symbol)

```


## 4. Choosing an ortholog filter

A common choice is to keep only relationships labelled `ortholog_one2one`. For
zebrafish, this label removes many genes that map unambiguously to human.

```{r the trap, echo = TRUE}

orth %>%
  dplyr::filter(zfish_symbol %in% c("serpinh1b", "col1a1a", "col1a1b", "hspa5", "col1a2")) %>%
  dplyr::select(zfish_symbol, human_symbol, orthology_type, n_human_partners)

```

`serpinh1b` is labelled `ortholog_one2many` but has `n_human_partners` equal to
1: it maps to a single human gene, `SERPINH1`. The "many" refers to the human
side, where `SERPINH1` has two zebrafish orthologs (`serpinh1a` and
`serpinh1b`). When mapping from zebrafish to human, this does not create
ambiguity.

`serpinh1b` is also one of the most significant up-regulated genes:

```{r serpinh1b, echo = TRUE}

up_ranked = de %>% dplyr::filter(log2FoldChange > 0) %>% dplyr::arrange(padj)

res %>% dplyr::filter(gene == "serpinh1b") %>% dplyr::select(gene, baseMean, log2FoldChange, padj)

# rank among the up-regulated genes by adjusted p-value
which(up_ranked$gene == "serpinh1b")

```

A filter on `ortholog_one2one` would remove it, together with `col1a1a`,
`col1a1b`, `col2a1a`, `ankrd1b` and other genes affected by the teleost
duplication.

This lesson therefore keeps zebrafish genes with exactly one human ortholog:

```{r unambiguous, echo = TRUE}

unambiguous = orth %>% filter(n_human_partners == 1)

to_human = function(fish_genes) {
  sort(unique(unambiguous$human_symbol[unambiguous$zfish_symbol %in% fish_genes]))
}

human_up   = to_human(genes_up)
human_down = to_human(genes_down)

length(genes_up);   length(human_up)
length(genes_down); length(human_down)

# The stored enrichment tables were produced from these exact lists. If Ensembl
# is updated and the mapping shifts, stop here rather than letting the counts
# quoted in the text disagree with the results shown below.
stopifnot(identical(human_up,   sort(readLines("../../data/enrichment/genes_up_human.txt"))),
          identical(human_down, sort(readLines("../../data/enrichment/genes_down_human.txt"))))

```

Different zebrafish genes can map to the same human gene (`col1a1a` and `col1a1b`
both map to `COL1A1`). Because only unique human symbols are kept, this can
reduce the size of the list but cannot inflate it. Section 6 shows the effect of
not applying the filter.


## 5. Enrichment with human annotation

```{r write human lists, echo = TRUE}

dir.create("outputs", showWarnings = FALSE)

writeLines(human_up,   "outputs/genes_up_human.txt")
writeLines(human_down, "outputs/genes_down_human.txt")

```

Submit both lists to **Enrichr** (<https://maayanlab.cloud/Enrichr/>) using
**GO Biological Process 2025**, as in S6. Save the up-regulated result as
`outputs/human_GO_up.txt` and the down-regulated result as
`outputs/human_GO_down.txt`. If the files are not present, the stored copies are
used.

```{r read human enrichment, echo = TRUE}

read_enrichment = function(mine, committed, submitted) {
  path = if (file.exists(mine)) mine else committed
  message("reading: ", path)

  # grouping_mark = "" stops a file that was opened and re-saved in a
  # comma-decimal spreadsheet from being read as numbers a million times too
  # large, which read_tsv would otherwise do without any warning.
  tb = read_tsv(path, show_col_types = FALSE,
                locale = locale(decimal_mark = ".", grouping_mark = ""))

  needed = c("Term", "Overlap", "Adjusted P-value", "Genes")
  missing = setdiff(needed, names(tb))
  if (length(missing) > 0) {
    stop(path, ": missing column(s) ", paste(missing, collapse = ", "),
         ". Is this an Enrichr export?", call. = FALSE)
  }
  if (nrow(tb) == 0) {
    stop(path, ": the file contains no rows.", call. = FALSE)
  }
  if (!is.numeric(tb$`Adjusted P-value`) || any(tb$`Adjusted P-value` > 1, na.rm = TRUE)) {
    stop(path, ": 'Adjusted P-value' is not a probability. If you opened the ",
         "file in a spreadsheet and saved it, download it again.", call. = FALSE)
  }

  # Check the table was produced from the list we think it was: a download
  # saved under the wrong name would otherwise be read without complaint.
  found = unique(unlist(strsplit(tb$Genes, ";")))
  shared = mean(toupper(found) %in% toupper(submitted))
  if (shared < 0.9) {
    stop(path, ": only ", round(100 * shared), "% of the genes in this table are ",
         "in the list submitted here. This looks like the wrong download.",
         call. = FALSE)
  }

  tb
}

# Rename each Enrichr download when saving it (see S6, section 5).
human_up_go = read_enrichment("outputs/human_GO_up.txt",
                              "../../data/enrichment/human_GO_BP_2025_up.tsv",
                              human_up)

human_up_go %>%
  dplyr::filter(`Adjusted P-value` < 0.05) %>%
  dplyr::select(Term, Overlap, `Adjusted P-value`, Genes) %>%
  dplyr::mutate(`Adjusted P-value` = formatC(`Adjusted P-value`, format = "e", digits = 2)) %>%
  as.data.frame()

```

With human annotation, the up-regulated list is enriched for **response to
unfolded protein**. The genes in this term are `DNAJA1`, `HSPA2`, `HSPA4`,
`HSP90AA1` and `HSPB1`, which are molecular chaperones, and `SERPINH1`. This
result was obtained from 131 human genes, fewer than the 354 zebrafish genes
tested in S6.

### Checking the mapping with the down-regulated list

A result that appears only after a transformation of the data should be checked
against a case where the expected outcome is known. The down-regulated list was
already enriched in the zebrafish analysis; after mapping, it should give a
similar result.

```{r human down, echo = TRUE}

human_down_go = read_enrichment("outputs/human_GO_down.txt",
                                "../../data/enrichment/human_GO_BP_2025_down.tsv",
                                human_down)

human_down_go %>%
  dplyr::filter(`Adjusted P-value` < 0.05) %>%
  dplyr::select(Term, Overlap, `Adjusted P-value`) %>%
  head(8) %>%
  dplyr::mutate(`Adjusted P-value` = formatC(`Adjusted P-value`, format = "e", digits = 2)) %>%
  as.data.frame()

```

```{r human down ecm genes, echo = TRUE}

# Genes in the extracellular matrix term, used again in section 8.
# Require exactly one significant match: GO term names change between releases
# (the 2021 library does not use this capitalisation), and a non-significant
# row would otherwise be used here without any indication.
ecm_row = human_down_go %>%
  dplyr::filter(grepl("extracellular matrix organization", Term, ignore.case = TRUE),
         `Adjusted P-value` < 0.05)

if (nrow(ecm_row) != 1) {
  stop("expected one significant extracellular matrix term, found ", nrow(ecm_row),
       call. = FALSE)
}

ecm_genes = unlist(strsplit(ecm_row$Genes, ";"))

ecm_genes

```

The human analysis again returns extracellular matrix and collagen fibril
organization, consistent with the zebrafish result. It also returns terms that
were not significant with the 2018 zebrafish library, such as
`Metal Ion Transport`.


## 6. Effect of not filtering orthologs

The following result was obtained by keeping all orthologs, including those where
one zebrafish gene maps to several human genes.

```{r allortho, echo = TRUE}

all_up_go = read_tsv("../../data/enrichment/human_GO_BP_2025_up_ALLORTHO.tsv",
                     show_col_types = FALSE)

all_up_go %>%
  dplyr::filter(`Adjusted P-value` < 0.05) %>%
  dplyr::select(Term, Overlap, `Adjusted P-value`) %>%
  head(5) %>%
  dplyr::mutate(`Adjusted P-value` = formatC(`Adjusted P-value`, format = "e", digits = 2))

# how many terms are significant here, against 2 with the filter applied
sum(all_up_go$`Adjusted P-value` < 0.05)

sum(human_up_go$`Adjusted P-value` < 0.05)

```

The top term is `Cellular Response to Zinc Ion` (adjusted p-value about 1.8e-19),
and 60 terms are significant instead of 2. The genes behind the top term:

```{r allortho genes, echo = TRUE}

all_up_go$Genes[1]

```

```{r mt2, echo = TRUE}

orth %>% dplyr::filter(zfish_symbol == "mt2") %>% dplyr::select(zfish_symbol, human_symbol, orthology_type)

```

Twelve of the 13 genes are human metallothioneins, and all twelve come from a
single zebrafish gene, `mt2`. One differentially expressed gene is counted twelve
times, which violates the assumption of the enrichment test that genes are
counted independently. The same occurs in the down-regulated list:

```{r ugt, echo = TRUE}

orth %>%
  dplyr::filter(zfish_symbol %in% genes_down, grepl("^ugt", zfish_symbol)) %>%
  dplyr::count(zfish_symbol, name = "human_partners")

```

When one gene maps to several orthologs, check how many human genes each
zebrafish gene contributes before interpreting the enrichment.


## 7. Effect of the annotation release

GO is updated regularly. The table below shows the same 131-gene list tested
against four releases of GO Biological Process.

```{r vintages, echo = TRUE}

vintages = c(2021, 2023, 2025, 2026)

summary_table = lapply(vintages, function(year) {
  tb = read_tsv(sprintf("../../data/enrichment/human_GO_BP_%d_up.tsv", year),
                show_col_types = FALSE)
  sig = tb %>% filter(`Adjusted P-value` < 0.05) %>% arrange(`Adjusted P-value`)
  data.frame(
    release     = year,
    terms_tested = nrow(tb),
    significant  = nrow(sig),
    # A release with no significant term would otherwise show as NA, which reads
    # like a formatting problem rather than a result.
    top_term     = if (nrow(sig) == 0) "none significant" else sub(" \\(GO:.*", "", sig$Term[1]),
    top_padj     = if (nrow(sig) == 0) NA_character_ else formatC(sig$`Adjusted P-value`[1], format = "e", digits = 2)
  )
})

do.call(rbind, summary_table)

```

With the same input, the number of significant terms is 18 in 2021, 10 in 2023,
2 in 2025 and 11 in 2026. The "response to unfolded protein" term in each release:

```{r upr across vintages, echo = TRUE}

upr = lapply(vintages, function(year) {
  tb = read_tsv(sprintf("../../data/enrichment/human_GO_BP_%d_up.tsv", year),
                show_col_types = FALSE)
  row = tb %>% filter(grepl("unfolded protein", Term, ignore.case = TRUE)) %>%
        arrange(`Adjusted P-value`) %>% head(1)
  # If a release names the term differently there is no matching row;
  # return NA instead of stopping the document.
  if (nrow(row) == 0) {
    data.frame(release = year, overlap = NA_character_, padj = NA_character_)
  } else {
    data.frame(release = year,
               overlap = row$Overlap,
               padj    = formatC(row$`Adjusted P-value`, format = "e", digits = 2))
  }
})

do.call(rbind, upr)

```

"Response to unfolded protein" is significant in all four releases, although the
number of genes annotated to the term changes between them. Most other terms are
significant in only some releases.

Results that remain significant across annotation releases are more reliable
than those that depend on a single release. Repeating an analysis with another
release requires changing only the library name.


## 8. Protein interaction networks: STRING

Enrichment tests whether a list is over-represented in annotation categories.
A network analysis instead examines known and predicted interactions between the
proteins in the list.

At <https://string-db.org>, choose **Multiple proteins**, paste the contents of
`outputs/genes_down_human.txt`, and select *Homo sapiens*.

The interactions are also stored as a table:

```{r string, echo = TRUE}

edges = read_tsv("../../data/enrichment/string_edges_human_down.tsv",
                 show_col_types = FALSE)

nrow(edges)

edges %>% dplyr::select(preferredName_A, preferredName_B, score) %>% head()

```

```{r string hubs, echo = TRUE}

# Proteins with the largest number of interactions
hubs = c(edges$preferredName_A, edges$preferredName_B) %>%
  table() %>%
  sort(decreasing = TRUE) %>%
  head(12)

hubs

```

**Question.** Compare the most connected proteins with `ecm_genes` from
section 5, the genes in "Extracellular Matrix Organization" for the same list.

```{r hubs vs ecm, echo = TRUE}

intersect(names(hubs), ecm_genes)

```

To what extent do the two approaches agree? Note that a highly connected protein
and a member of an enriched term are identified by different criteria.


## 9. Pathway analysis: Reactome

Reactome is a manually curated database of human pathways.

Paste `outputs/genes_down_human.txt` into <https://reactome.org> → *Analyse gene
list*, or read the stored result:

```{r reactome, echo = TRUE}

reactome = read_tsv("../../data/enrichment/human_Reactome_2024_down.tsv",
                    show_col_types = FALSE)

reactome %>%
  dplyr::filter(`Adjusted P-value` < 0.05) %>%
  dplyr::select(Term, Overlap, `Adjusted P-value`) %>%
  head(10) %>%
  dplyr::mutate(`Adjusted P-value` = formatC(`Adjusted P-value`, format = "e", digits = 2))

```

**Question.** Where do the Reactome and GO results agree, and which results
appear in only one of them?


## 10. Exercise: interpreting the results together

The analyses above show, for the same experiment:

- collagen and extracellular matrix genes are down-regulated
- chaperones and endoplasmic reticulum folding genes are up-regulated
- `serpinh1b` (human `SERPINH1`) is up-regulated about thirteen-fold

```{r serpinh1 identity, echo = TRUE}

res %>%
  dplyr::filter(gene %in% c("serpinh1b", "hspa5", "hsp90b1", "pdia4",
                     "col1a1a", "col1a2", "col11a2")) %>%
  dplyr::select(gene, baseMean, log2FoldChange, padj) %>%
  dplyr::arrange(log2FoldChange)

```

**Question.** `SERPINH1`, also known as HSP47, is a collagen-specific molecular
chaperone in the endoplasmic reticulum.

Propose an explanation that connects the three observations above, and describe
an experiment that could test it.

**Question.** The human analysis used 131 of the 354 up-regulated zebrafish
genes. What happened to the remaining 223, and how could their exclusion affect
the result?


## 11. Print R session info

```{r print out R session info, echo = TRUE}
# Always good practice to print out the R session info for reproducibility
sessionInfo()
```
