---
title: "S6_Gene_Lists_to_Biology"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Day 1 ended with a volcano plot: a few hundred genes that changed between mutant
and wild type. That is a list of names. It is not yet a statement about biology.

This lesson turns the list into a statement, and then teaches you to distrust it.


## 1. Load libraries

```{r load libraries, echo = TRUE, message=FALSE, warning=FALSE}

library(dplyr, quietly = TRUE)
library(readr, quietly = TRUE)
library(ggplot2, quietly = TRUE)

```


## 2. Where day 1 left us

```{r load results, echo = TRUE}

# The results table written by S4, and visualised in S5.
res = read_tsv("../../results/DESeq2_results.tsv", show_col_types = FALSE)

head(res)

nrow(res)

```


## 3. Build the gene lists

An enrichment tool takes a *set* of genes, not a ranked table. So the first
decision is which genes get in.

```{r build lists, echo = TRUE}

# padj  : how confident are we that this gene changed at all
# log2FoldChange : how much did it change
#
# Both matter. A gene can be significant and barely move; a gene can move a lot
# and be noise.
padj_max = 0.05
lfc_min  = 1

de = res %>%
  filter(!is.na(padj), padj < padj_max, abs(log2FoldChange) > lfc_min)

genes_up   = sort(de$gene[de$log2FoldChange > 0])
genes_down = sort(de$gene[de$log2FoldChange < 0])

length(genes_up)
length(genes_down)

```

Notice that these cutoffs are **not** the ones the volcano plot used in S5
(`padj < 0.001`, `|log2FC| > 2.5`, which keep only 48 and 55 genes).

That was not an inconsistency. The volcano's cutoffs exist so the gene labels
stay readable on a crowded figure. Choosing genes for enrichment is a different
job, and it gets its own decision. Be suspicious of any analysis that reuses a
threshold without saying why.

```{r save lists, echo = TRUE}

# Save the lists so we can paste them into a website in a moment.
dir.create("outputs", showWarnings = FALSE)

writeLines(genes_up,   "outputs/genes_up.txt")
writeLines(genes_down, "outputs/genes_down.txt")

# What we are about to hand over is just names, one per line:
head(genes_down, 10)

```


## 4. Where a gene list can be taken

A gene symbol is a key into a large number of public databases. Knowing which
one answers which question is most of the skill.

| Resource | Answers |
| --- | --- |
| **NCBI Gene** | What is this gene, in any organism? |
| **Ensembl** | Where is it in the genome, and what is its equivalent in another species? |
| **ZFIN** | Everything zebrafish: expression patterns, mutants, phenotypes |
| **UniProt** | What does the protein do, and what does it look like? |
| **GO** | Controlled vocabulary of biological processes, functions and locations |
| **KEGG** / **Reactome** | Curated pathway maps |
| **STRING** | Which proteins interact with which |
| **Enrichr** | Is my gene list unusually full of any of the above? |

The one we need now is the last. We have 588 down-regulated genes; we want to
know whether they share anything beyond being on the same list.


## 5. Enrichment: the idea

Enrichment asks one question, repeated thousands of times:

> Of the 588 genes on my list, 21 are annotated "extracellular matrix
> organization". The genome has 92 such genes in total. Is 21 more than I would
> expect by chance?

That is a hypergeometric test, the same mathematics as drawing coloured balls
from an urn. Doing it for every GO term at once is why the output has an
adjusted p-value column: thousands of tests need a multiple-testing correction,
exactly as in S4.

### Run it on the website

FishEnrichr is the zebrafish version of Enrichr.

1. Open <https://maayanlab.cloud/FishEnrichr/>
2. Open `outputs/genes_down.txt`, copy all of it, paste it into the box
3. Submit, then open the **GO Biological Process 2018** results
4. Use **Export entries to table** to download the table
5. Save it into this lesson's `outputs/` folder

Then do the same for `outputs/genes_up.txt`.

### Or run it from R

```{r enrichr from R, echo = TRUE, eval = FALSE}

# The enrichR package talks to the same service.
library(enrichR)

setEnrichrSite("FishEnrichr")

enriched = enrichr(genes_down, "GO_Biological_Process_2018")

head(enriched[["GO_Biological_Process_2018"]])

```

Either route reaches the same server, and the server is not always available.
If a whole classroom submits at once you will meet `HTTP 429 Too Many Requests`.
That is normal, and it is why the next chunk does not depend on your download
succeeding.


## 6. Read the results back into R

```{r read enrichment, echo = TRUE}

# Use your own download if you made one; otherwise the copy committed with the
# course. Both are the same file.
read_enrichment = function(mine, committed) {
  path = if (file.exists(mine)) mine else committed
  message("reading: ", path)
  read_tsv(path, show_col_types = FALSE)
}

fish_down = read_enrichment("outputs/GO_Biological_Process_2018_table.txt",
                            "../../data/enrichment/fish_GO_BP_2018_down.tsv")

# One row per GO term tested.
dim(fish_down)

names(fish_down)

```

```{r significant terms, echo = TRUE}

sig_down = fish_down %>%
  filter(`Adjusted P-value` < 0.05) %>%
  arrange(`Adjusted P-value`)

sig_down %>% select(Term, Overlap, `Adjusted P-value`)

```

Five terms out of 918 tested. Now the up-regulated genes:

```{r up enrichment, echo = TRUE}

fish_up = read_enrichment("outputs/GO_Biological_Process_2018_table_up.txt",
                          "../../data/enrichment/fish_GO_BP_2018_up.tsv")

sum(fish_up$`Adjusted P-value` < 0.05)

```

Zero. Three hundred and fifty-four genes go up, strongly and significantly, and
they enrich in nothing at all.

Hold on to that. S7 is about what it means.


## 7. Plot the enrichment

The downloaded table is an ordinary data frame, so everything from S3 applies.

```{r enrichment dotplot, echo = TRUE, fig.width = 9, fig.height = 4}

# "Overlap" arrives as the text "21/92". Split it to get a gene ratio.
plot_data = sig_down %>%
  mutate(
    n_genes   = as.numeric(sub("/.*", "", Overlap)),
    term_size = as.numeric(sub(".*/", "", Overlap)),
    ratio     = n_genes / term_size,
    Term      = sub(" \\(GO:.*", "", Term)          # drop the GO id for the label
  )

ggplot(plot_data, aes(x = ratio, y = reorder(Term, ratio))) +
  geom_point(aes(size = n_genes, colour = `Adjusted P-value`)) +
  scale_colour_gradient(low = "#b2182b", high = "#2166ac") +
  labs(x = "fraction of the GO term found in our list",
       y = NULL,
       size = "genes",
       title = "Down-regulated genes, GO Biological Process (zebrafish, 2018)") +
  theme_bw()

```

This is the figure that goes in a talk. It is also the point at which most
people stop, and that is a mistake.


## 8. Read the genes, not the term

A GO term is a label attached by a curator, sometimes decades ago, sometimes by
an automatic pipeline. The label can be misleading even when the statistics are
correct. The `Genes` column is the evidence, and it takes ten seconds to read.

```{r driver genes, echo = TRUE}

sig_down %>%
  select(Term, Overlap, Genes) %>%
  mutate(Term = sub(" \\(GO:.*", "", Term)) %>%
  as.data.frame()

```

Look at those five rows carefully before reading on.

Three things should bother you.

**The top term is one gene family.** All 21 genes behind "extracellular matrix
organization" are collagens, plus one collagen cross-linking enzyme. That is a
real and coherent finding, but it is one story, not 21 independent ones.

**Three terms are the same genes wearing different hats.** "Notochord
morphogenesis" and "skeletal system development" are driven by the same
collagens as the term above them. Ranked lists make these look like three
findings. They are one.

**One term is not about what it says.** "Cellular response to radiation" is
carried by `opn1sw1`, `opn1sw2`, `rhol`, `rho`, `opn4a`, `opn1lw2`, `opn1mw1` —
seven opsins and a rhodopsin. There is no radiation in this experiment. These
are the same photoreceptor genes that drive "response to light stimulus", and
they are annotated to a radiation term because light is electromagnetic
radiation.

So the rule, which is worth more than any of the code above:

> **Never report a GO term before reading the genes that produced it.**


## 9. Exercise: why are photoreceptor genes here at all?

These are supposed to be dissected hearts at 48 hpf.

```{r photoreceptor check, echo = TRUE}

res %>%
  filter(gene %in% c("rho", "crx", "opn1sw1", "opn1sw2", "opn1lw2", "opn1mw1")) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(desc(baseMean))

```

`baseMean` is the average expression across all samples. Rhodopsin sits at
around 2,100 and `crx`, the master regulator of photoreceptor identity, at
around 2,900.

**Question.** Is this a discovery about the mutant, or a problem with the
experiment? What would you check next?

**Now try the same for these genes**, and decide whether the answer is the same
kind of problem:

```{r circadian check, echo = TRUE}

res %>%
  filter(gene %in% c("per1a", "per2", "per3", "nr1d1", "npas2", "cry5")) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(log2FoldChange)

```

Both sets of genes are real, strong and statistically significant. Neither is
about the mutation. One tells you what tissue is in the tube; the other tells
you something about when the samples were collected. Finding them is not a
failure of the analysis — it is the analysis working.


## 10. Exercise: a curated list of your own

`data/chd_genes.annotations.tsv` holds 276 genes associated with congenital
heart disease. Our mutant is a heart model, so those genes ought to be
over-represented among the differentially expressed ones.

```{r chd overlap, echo = TRUE}

chd = read_tsv("../../data/chd_genes.annotations.tsv", show_col_types = FALSE)

# The CHD list uses human symbols (upper case); ours are zebrafish (lower case).
chd_symbols = tolower(chd$gene_symbol)

in_chd = tolower(res$gene) %in% chd_symbols
is_de  = res$gene %in% de$gene

table(CHD = in_chd, DE = is_de)

fisher.test(table(in_chd, is_de))$p.value

```

**Question.** The p-value is not significant. Does that mean the CHD gene list
is irrelevant to this experiment? Look at *which* CHD genes are differentially
expressed before you answer:

```{r chd genes de, echo = TRUE}

res %>%
  filter(tolower(gene) %in% chd_symbols, gene %in% de$gene) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(log2FoldChange)

```

One more thing to notice, because S7 begins here. Lower-casing a human symbol to
match a zebrafish one is a crude way to connect two species. It finds `col1a2`,
but it misses `col1a1a` and `col1a1b` — the two zebrafish copies of human
`COL1A1` — because zebrafish duplicated its genome and renamed the copies.

Matching species properly is the subject of the next lesson.


## 11. Print R session info

```{r print out R session info, echo = TRUE}
# Always good practice to print out the R session info for reproducibility
sessionInfo()
```
