---
title: "S6_Gene_Lists_to_Biology"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Day 1 ended with a volcano plot: a few hundred genes that changed between mutant
and wild type. That is a list of names. It is not yet a statement about biology.

This lesson turns the list into a statement, and then teaches you to distrust it.


## 1. Load libraries

```{r load libraries, echo = TRUE, message=FALSE, warning=FALSE}

library(dplyr, quietly = TRUE)
library(readr, quietly = TRUE)
library(ggplot2, quietly = TRUE)

```


## 2. Where day 1 left us

```{r load results, echo = TRUE}

# The results table written by S4, and visualised in S5.
res = read_tsv("../../results/DESeq2_results.tsv", show_col_types = FALSE)

head(res)

nrow(res)

```


## 3. Build the gene lists

An enrichment tool takes a *set* of genes, not a ranked table. So the first
decision is which genes get in.

```{r build lists, echo = TRUE}

# padj  : how confident are we that this gene changed at all
# log2FoldChange : how much did it change
#
# Both matter. A gene can be significant and barely move; a gene can move a lot
# and be noise.
padj_max = 0.05
lfc_min  = 1

de = res %>%
  filter(!is.na(padj), padj < padj_max, abs(log2FoldChange) > lfc_min)

genes_up   = sort(de$gene[de$log2FoldChange > 0])
genes_down = sort(de$gene[de$log2FoldChange < 0])

length(genes_up)
length(genes_down)

```

Notice that these cutoffs are **not** the ones the volcano plot used in S5.
That plot passed `y = 'pvalue'`, so it highlighted genes with a raw
`pvalue < 0.001` and `|log2FC| > 2.5` — 55 up and 56 down, 111 points in all.

That was not an inconsistency. The volcano's cutoffs exist so the gene labels
stay readable on a crowded figure. Choosing genes for enrichment is a different
job, and it gets its own decision. Be suspicious of any analysis that reuses a
threshold without saying why.

```{r save lists, echo = TRUE}

# Save the lists so we can paste them into a website in a moment.
dir.create("outputs", showWarnings = FALSE)

writeLines(genes_up,   "outputs/genes_up.txt")
writeLines(genes_down, "outputs/genes_down.txt")

# What we are about to hand over is just names, one per line:
head(genes_down, 10)

```


## 4. Where a gene list can be taken

A gene symbol is a key into a large number of public databases. Knowing which
one answers which question is most of the skill.

| Resource | Answers |
| --- | --- |
| **NCBI Gene** | What is this gene, in any organism? |
| **Ensembl** | Where is it in the genome, and what is its equivalent in another species? |
| **ZFIN** | Everything zebrafish: expression patterns, mutants, phenotypes |
| **UniProt** | What does the protein do, and what does it look like? |
| **GO** | Controlled vocabulary of biological processes, functions and locations |
| **KEGG** / **Reactome** | Curated pathway maps |
| **STRING** | Which proteins interact with which |
| **Enrichr** | Is my gene list unusually full of any of the above? |

The one we need now is the last. We have 588 down-regulated genes; we want to
know whether they share anything beyond being on the same list.


## 5. Enrichment: the idea

Enrichment asks one question, repeated thousands of times:

> Of the 588 genes on my list, 21 are annotated "extracellular matrix
> organization". The genome has 92 such genes in total. Is 21 more than I would
> expect by chance?

That is a hypergeometric test, the same mathematics as drawing coloured balls
from an urn. Doing it for every GO term at once is why the output has an
adjusted p-value column: thousands of tests need a multiple-testing correction,
exactly as in S4.

### Run it on the website

FishEnrichr is the zebrafish version of Enrichr.

1. Open <https://maayanlab.cloud/FishEnrichr/>
2. Open `outputs/genes_down.txt`, copy all of it, paste it into the box
3. Submit, then open the **GO Biological Process 2018** results
4. Use **Export entries to table** to download the table
5. Save it into this lesson's `outputs/` folder, renaming it to
   **`fish_GO_down.txt`**

Then do the same for `outputs/genes_up.txt`, saving that one as
**`fish_GO_up.txt`**.

The rename matters: Enrichr gives both downloads the same default filename, so
without it the second one overwrites the first and the chunks below silently
read the wrong list.

### Or run it from R

```{r enrichr from R, echo = TRUE, eval = FALSE}

# The enrichR package talks to the same service.
library(enrichR)

setEnrichrSite("FishEnrichr")

enriched = enrichr(genes_down, "GO_Biological_Process_2018")

head(enriched[["GO_Biological_Process_2018"]])

```

Either route reaches the same server, and the server is not always available.
If a whole classroom submits at once you will meet `HTTP 429 Too Many Requests`.
That is normal, and it is why the next chunk does not depend on your download
succeeding.


## 6. Read the results back into R

```{r read enrichment, echo = TRUE}

# Use your own download if you made one; otherwise the copy committed with the
# course. Both are the same file.
read_enrichment = function(mine, committed) {
  path = if (file.exists(mine)) mine else committed
  message("reading: ", path)
  read_tsv(path, show_col_types = FALSE)
}

fish_down = read_enrichment("outputs/fish_GO_down.txt",
                            "../../data/enrichment/fish_GO_BP_2018_down.tsv")

# One row per GO term tested.
dim(fish_down)

names(fish_down)

```

```{r significant terms, echo = TRUE}

sig_down = fish_down %>%
  filter(`Adjusted P-value` < 0.05) %>%
  arrange(`Adjusted P-value`)

sig_down %>% select(Term, Overlap, `Adjusted P-value`)

```

Five terms out of 918 tested. Now the up-regulated genes:

```{r up enrichment, echo = TRUE}

fish_up = read_enrichment("outputs/fish_GO_up.txt",
                          "../../data/enrichment/fish_GO_BP_2018_up.tsv")

sum(fish_up$`Adjusted P-value` < 0.05)

```

Zero. Three hundred and fifty-four genes go up, strongly and significantly, and
they enrich in nothing at all.

Hold on to that. S7 is about what it means.


## 7. Plot the enrichment

The downloaded table is an ordinary data frame, so everything from S3 applies.

```{r enrichment dotplot, echo = TRUE, fig.width = 9, fig.height = 4}

# "Overlap" arrives as the text "21/92". Split it to get a gene ratio.
plot_data = sig_down %>%
  mutate(
    n_genes   = as.numeric(sub("/.*", "", Overlap)),
    term_size = as.numeric(sub(".*/", "", Overlap)),
    ratio     = n_genes / term_size,
    # p-values span eight orders of magnitude; on a linear colour ramp every
    # strong term would get the same colour and only the weakest would stand out.
    neg_log10_padj = -log10(`Adjusted P-value`),
    Term      = sub(" \\(GO:.*", "", Term)          # drop the GO id for the label
  )

ggplot(plot_data, aes(x = ratio, y = reorder(Term, ratio))) +
  geom_point(aes(size = n_genes, colour = neg_log10_padj)) +
  scale_colour_gradient(low = "#92c5de", high = "#b2182b") +
  labs(x = "fraction of the GO term found in our list",
       y = NULL,
       size = "genes",
       colour = "-log10(padj)",
       title = "Down-regulated genes, GO Biological Process (zebrafish, 2018)") +
  theme_bw()

```

This is the figure that goes in a talk. It is also the point at which most
people stop, and that is a mistake.


## 8. Read the genes, not the term

A GO term is a label attached by a curator, sometimes decades ago, sometimes by
an automatic pipeline. The label can be misleading even when the statistics are
correct. The `Genes` column is the evidence, and it takes ten seconds to read.

```{r driver genes, echo = TRUE}

sig_down %>%
  select(Term, Overlap, Genes) %>%
  mutate(Term = sub(" \\(GO:.*", "", Term)) %>%
  as.data.frame()

```

Look at those five rows carefully before reading on.

Three things should bother you.

**The top term is mostly one gene family.** Nineteen of the 21 genes behind
"extracellular matrix organization" are collagens. The other two are `loxl3a`, a
collagen cross-linking enzyme, and `zgc:113232`, an unnamed clone identifier.
That is a real and coherent finding, but it is largely one story about collagen,
not 21 independent ones.

**Other terms re-use those same collagens.** "Notochord morphogenesis" shares 3
of its 6 genes with the term above it, and "skeletal system development" shares 7
of its 18. So they are not independent confirmations — but they are not pure
duplicates either. Count the shared genes before deciding how much a second term
adds; the ranked list will not do it for you.

**One term is not about what it says.** "Cellular response to radiation" is
carried by `opn1sw1`, `opn1sw2`, `opn1lw2`, `opn1mw1`, `opn4a`, `rho`, `rhol`
and `bbc3` — five cone opsins, two rhodopsins and one other gene. There was no
radiation in this experiment. The seven photoreceptor genes are annotated here
because light *is* electromagnetic radiation. They also sit inside "response to
light stimulus" — but that term has 14 genes, and five of the others are clock
genes (`per1a`, `per2`, `per3`, `cry5`, `npas2`). So "response to light
stimulus" mixes two separate problems that section 9 pulls apart.

The eighth, `bbc3`, is PUMA, and it genuinely is a DNA-damage response gene. One
gene out of eight, arriving with seven that do not belong, is not evidence of
anything. Noticing that it is the only honest member of the set is exactly the
kind of reading this section is asking for.

So the rule, which is worth more than any of the code above:

> **Never report a GO term before reading the genes that produced it.**


## 9. Exercise: why are photoreceptor genes here at all?

These are supposed to be dissected hearts at 48 hpf.

```{r photoreceptor check, echo = TRUE}

res %>%
  filter(gene %in% c("rho", "crx", "opn1sw1", "opn1sw2", "opn1lw2", "opn1mw1")) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(desc(baseMean))

```

`baseMean` is the average expression across all samples. Rhodopsin sits at
around 2,100 and `crx`, the master regulator of photoreceptor identity, at
around 2,900.

**Question.** Is this a discovery about the mutant, or a problem with the
experiment? What would you check next?

**Now try the same for these genes**, and decide whether the answer is the same
kind of problem:

```{r circadian check, echo = TRUE}

res %>%
  filter(gene %in% c("per1a", "per2", "per3", "nr1d1", "npas2", "cry5")) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(log2FoldChange)

```

Both sets of genes are real, strong and statistically significant. Neither is
about the mutation. One tells you what tissue is in the tube; the other tells
you something about when the samples were collected. Finding them is not a
failure of the analysis — it is the analysis working.


## 10. Exercise: a curated list of your own

`data/chd_genes.annotations.tsv` holds 276 genes associated with congenital
heart disease. Our mutant is a heart model, so those genes ought to be
over-represented among the differentially expressed ones.

```{r chd overlap, echo = TRUE}

chd = read_tsv("../../data/chd_genes.annotations.tsv", show_col_types = FALSE)

# The CHD list uses human symbols (upper case); ours are zebrafish (lower case).
chd_symbols = tolower(chd$gene_symbol)

in_chd = tolower(res$gene) %in% chd_symbols
is_de  = res$gene %in% de$gene

table(CHD = in_chd, DE = is_de)

fisher.test(table(in_chd, is_de))$p.value

```

**Question.** The p-value is not significant. Does that mean the CHD gene list
is irrelevant to this experiment? Look at *which* CHD genes are differentially
expressed before you answer:

```{r chd genes de, echo = TRUE}

res %>%
  filter(tolower(gene) %in% chd_symbols, gene %in% de$gene) %>%
  select(gene, baseMean, log2FoldChange, padj) %>%
  arrange(log2FoldChange)

```

One more thing to notice, because S7 begins here. Lower-casing a human symbol to
match a zebrafish one is a crude way to connect two species. It finds `col1a2`,
but it misses `col1a1a` and `col1a1b` — the two zebrafish copies of human
`COL1A1` — because zebrafish duplicated its genome and renamed the copies.

Matching species properly is the subject of the next lesson.


## 11. Print R session info

```{r print out R session info, echo = TRUE}
# Always good practice to print out the R session info for reproducibility
sessionInfo()
```
