---
title: "S2_IO_in_R"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import/export data into/from R

# 0. Load libraries

```{r Load libraries, echo=TRUE}

library(readxl) 
library(microbenchmark)

```

## 1. Import data

```{r Import data, echo=TRUE}

# 1. Import data

# Import data from a csv file
df <- read.csv("../../data/chd_genes.annotations.csv", 
               sep = ",", 
               header = TRUE)

# print the head of the data frame
head(df)

# Import data using the read.table function
df <- read.table("../../data/chd_genes.annotations.tsv",
                 sep = "\t", 
                 header = TRUE)

# print the head of the data frame
head(df)

# Import data using the read.delim function
df <- read.delim("../../data/chd_genes.annotations.tsv",
                 sep = "\t", 
                 header = TRUE)

# print the head of the data frame
head(df)

# End of the section


```

\newpage

## 2. Export data

```{r Export data, echo=TRUE}

# 2. Export data

# load Iris data set
data(iris)

# print the head of the data frame
head(iris)

# Export data using the write.csv function
write.csv(iris, file = "iris.csv")

# Export data using the write.table function
write.table(iris, file = "iris.txt")

# End of the section


```


\newpage

## 3. Import data from Excel

```{r Import_data_from_Excel, echo=TRUE}

# 3. Import data from Excel

# Import data from Excel
df <- read_excel("../../data/chd_genes.annotations.xlsx")

# print the head of the data frame
head(df)

# End of the section


```

\newpage

## 4. Evaluating speed performance of import functions.

Note: Differences between <read.csv>, <read.table> and <read.delim>
read.csv is a special case of read.table, which is a special case of read.delim.
read.csv uses a comma as a separator, read.table uses a tab, and read.delim uses a tab.
read.csv is the most common of the three, and read.table is the most flexible.
read.delim is used when the data is tab-delimited, but the extension is not txt.

```{r Bonus, fig.height=6, fig.width=6, message=TRUE, paged.print=TRUE}

# What is the faster function to import data?
# read.csv is faster than read.table and read.delim.

# Import data using the read.csv function
microbenchmark(
  read.csv("../../data/chd_genes.annotations.tsv",
                        sep = "\t", 
                        header = TRUE)
  )

# Import data using the read.table function
microbenchmark(
  read.table("../../data/chd_genes.annotations.tsv",
                          sep = "\t", 
                          header = TRUE)
  )

# Import data using the read.delim function
microbenchmark(
  read.delim("../../data/chd_genes.annotations.tsv",
                          sep = "\t", 
                          header = TRUE)
  )


# End of the section
```


