---
title: "S1_R_Data_Structures"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output: 
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introcution to R data structures

## 1. Vectors

A vector is a collection of elements of the same data type. The elements of a vector 
are indexed from 1 to the length of the vector. The elements of a vector can be 
accessed using the index of the element.


```{r Vectors, echo=TRUE}

# Vectors

# Create a vector
v <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# Print the vector
v

# Print the class of the vector
class(v)

# Print the length of the vector
length(v)

# Print the first element of the vector
v[1]

# Print the last element of the vector
v[length(v)]

# Print the first 5 elements of the vector
v[1:5]

```


## 2. Factors

A factor is a vector that can contain only predefined values, and is used to store 
categorical data. The predefined values are called levels. The levels of a factor 
are ordered alphabetically by default. The levels of a factor can be ordered in a different way.


```{r Factors, echo=TRUE}

# Factors

# Create a factor
fa <- factor(c("a", "b", "c", "a", "b", "c"))

# Print the factor
fa

# Print the class of the factor
class(fa)

# Print the levels of the factor
levels(fa)

# Print the number of levels of the factor
nlevels(fa)

# Print the number of elements of the factor
length(fa)

# reoder the levels of the factor
fa <- factor(fa, levels = c("c", "b", "a"))

# Print the levels of the factor
levels(fa)


# End of the section


```


## 3. Matrices

A matrix is a collection of elements of the same data type arranged in a two-dimensional rectangular layout.
The elements of a matrix are indexed from 1 to the number of rows and columns of the matrix. The elements 
of a matrix can be accessed using the index of the element.


```{r Matrices, echo=TRUE}

# Matrices

# Create a matrix
x <- matrix(1:12, nrow = 3, ncol = 4)

# Print the matrix
x

# Print the class of the matrix
class(x)

# Print the number of rows and columns of the matrix
dim(x)

# Print the first element of the matrix
x[1, 1]

# Print the last element of the matrix
x[3, 4]

# Print the first row of the matrix
x[1, ]

# Print the first column of the matrix
x[, 1]

# Explain the difference between a vector and a matrix
# A vector is a collection of elements of the same data type arranged in a one-dimensional layout.
# A matrix is a collection of elements of the same data type arranged in a two-dimensional rectangular layout.

```


### 3.1 Operations with matrices

```{r Operations with matrices, echo=TRUE}

# Operations with matrices

# Multiplication of matrices
# The number of columns of the first matrix must be equal to the number of rows of the second matrix.
# The result of the multiplication of two matrices is a matrix with the same number of rows as the first
# matrix and the same number of columns as the second matrix.

# Create two matrices
x <- matrix(1:12, nrow = 3, ncol = 4)
y <- matrix(1:12, nrow = 4, ncol = 3)

# Multiply the two matrices
x %*% y

# Transpose of a matrix
# The transpose of a matrix is a matrix whose rows are the columns of the original matrix and 
# whose columns are the rows of the original matrix.

# Transpose the matrix x
t(x)

# Other matrix operations: rowSums, colSums, rowMeans, colMeans

# rowSums
# The rowSums function computes the sum of the elements of each row of a matrix.
result <- rowSums(x)
result

# colSums
# The colSums function computes the sum of the elements of each column of a matrix.
result <- colSums(x)
result

# rowMeans
# The rowMeans function computes the mean of the elements of each row of a matrix.
result <- rowMeans(x)
result

# colMeans
# The colMeans function computes the mean of the elements of each column of a matrix.
result <- colMeans(x)
result

# End of the section

```

## 4. Arrays

An array is a collection of elements of the same data type arranged in a multi-dimensional rectangular layout.
In R, an array can have any number of dimensions.
For example:
A one-dimensional array is a vector.
A two-dimensional array is a matrix.
A three-dimensional array is a collection of matrices.
The elements of an array are indexed from 1 to the length of the array in each dimension.


```{r Arrays, echo=TRUE}

# Arrays

# Create an array
a <- array(1:24, dim = c(3, 4, 2))

# Print the array
a

# Print the class of the array
class(a)

# Print the number of rows, columns and layers of the array
dim(a)

# Print the first element of the array
a[1, 1, 1]

# Print the last element of the array
a[3, 4, 2]

# Print the first row of the array
a[1, , ]

# Print the first column of the array
a[, 1, ]

# Print the first layer of the array
a[,, 1]

# End of the section


```


## 5. Lists

A list is a collection of elements of different data types. The elements of 
a list are indexed from 1 to the length of the list.

```{r Lists, echo=TRUE}

# Lists

# Create a list
l <- list(1:10, "a", TRUE, 1.5)

# Print the list
l

# Print the class of the list
class(l)

# Print the length of the list
length(l)

# Print the first element of the list
l[[1]]

# Print the last element of the list
l[[length(l)]]

# Print the first 3 elements of the list
l[1:3]

# Differences between two [[ ]] and one [ ]:
# x[[1]] returns the first element of the list.
# x[1] returns a list with the first element of the list.


# End of the section
```

### 5.1 Named lists

```{r Named lists, echo=TRUE}
  
# Named lists

# Create a named list
l <- list(a = 1:10, b = "a", c = TRUE, d = 1.5)

# Print the list
l

# Print the class of the list
class(l)

# Print the length of the list
length(l)

# Print the first element of the list
l[[1]]

# Print the last element of the list
l[[length(l)]]

# Print the first 3 elements of the list
l[1:3]

# Print the element of the list with the name "a"
l[["a"]]

# Print the element of the list with the name "b"
l[["b"]]

# Print the element of the list with the name "c"
l[["c"]]

# Print the element of the list with the name "d"
l[["d"]]

# End of the section

```


## 6. Data frames

A data frame is a collection of elements of different data types arranged in a two-dimensional 
rectangular layout. The elements of a data frame are indexed from 1 to the length of the data 
frame in each dimension. The columns of a data frame can be of different data types.


```{r Data frames, echo=TRUE}


# Create a data frame
df <- data.frame(a = 1:10, b = "a", c = TRUE, d = 1.5)

# Print the data frame
df

# Print the head of the data frame
head(df)

# Print the tail of the data frame
tail(df)

# Print the class of the data frame
class(df)

# Print the number of rows and columns of the data frame
dim(df)

# Print the number of rows of the data frame
nrow(df)

# Print the number of columns of the data frame
ncol(df)

# End of the section


```

### 6.1 Operations on dataframes

```{r Operations on dataframes, echo=TRUE}

# 1. Filtering dataframes

# load the iris dataset
data(iris)


# Filter dataframes using the subset function
# Filter the dataframe to get the rows where the Sepal.Length is greater than 5
# and the Sepal.Width is greater than 3.5
df <- subset(iris, Sepal.Length > 5 & Sepal.Width > 3.5)

# print the head of the dataframe
head(df)

# Filter dataframes using the [] operator
# Filter the dataframe to get the rows where the Sepal.Length is greater than 5
# and the Sepal.Width is greater than 3.5

df <- iris[iris$Sepal.Length > 5 & iris$Sepal.Width > 3.5, ]

# print the head of the dataframe
head(df)

# 2. Aggregating dataframes using the dplyr package

# load the iris dataset
data(iris)

# load the dplyr package
library(dplyr)

# Group the dataframe by Species and get the mean of the Sepal.Length
df <- iris %>% group_by(Species) %>% summarise(mean(Sepal.Length))

# print the dataframe
df

# Group the dataframe by Species and get the mean of the Sepal.Length and Sepal.Width
df <- iris %>% group_by(Species) %>% summarise(mean(Sepal.Length), mean(Sepal.Width))

# print the dataframe
df

# End of the section

```
