---
title: "Installation and setup — R track"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output:
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## What this course is pinned to

| | |
|---|---|
| R | **4.6.x** |
| Bioconductor | **3.23** |
| CRAN snapshot | **2026-09-01** |

Bioconductor 3.23 requires R 4.6, so these move together or not at all.

Install R 4.6.x from [cran.r-project.org](https://cran.r-project.org) and RStudio
Desktop from [posit.co](https://posit.co/download/rstudio-desktop/). If you already
have an older R, install 4.6 alongside it and pick it in RStudio under
**Tools → Global Options → General → R version**.

## Install everything

Open `ZebraQ.Rproj` in RStudio, then run this once:

```{r install}
source("setup/install_r.R")
```

That is the whole setup. The script checks your R version, installs the nine CRAN
packages and two Bioconductor packages the lessons use, and prints the version of
each so you can see that it worked.

## Why a script instead of `install.packages()`

Two problems come up on people's own laptops, and the script exists to avoid both.

**Everyone ends up with different versions.** A plain `install.packages("ggplot2")`
gives you whatever CRAN published today. Over a semester the course drifts apart
from itself, and a plot that works on one laptop looks different on another. The
script installs from a *dated snapshot* of CRAN, so everybody gets the same
versions no matter when they run it.

**R offers to compile from source, and compiling fails.** On macOS and Windows the
default `pkgType` is `"both"`, which means R asks

> Do you want to install from sources the package which needs compilation?

whenever the source version is newer than the available binary. Saying yes needs a
compiler toolchain — Rtools on Windows, the Xcode command line tools on macOS —
which most people do not have, and the install dies partway through. The script
sets `pkgType = "binary"` so the question never comes up.

The packages come from [Posit Package Manager](https://packagemanager.posit.co),
which serves prebuilt binaries for Windows, macOS (both Intel and Apple Silicon)
and Linux. Plain CRAN has no Linux binaries at all, so a Linux install without this
would compile every package from source.

## If something goes wrong

**"This course is pinned to R 4.6 … you are running R 4.x"** — the script stopped
on purpose. Install R 4.6.x and select it in RStudio as described above.

**A package still fails.** Run the script again; it only installs what is missing,
so it is safe to repeat. If the same package fails twice, send the last 20 lines of
output rather than a screenshot of the red text.

**You want to know exactly what you have.** `sessionInfo()` prints your R version
and every loaded package — the R equivalent of the `session_info.show()` you will
see at the end of each Python notebook.

## Packages this installs

From CRAN: `ggplot2`, `ggpubr`, `dplyr`, `RColorBrewer`, `readxl`, `openxlsx`,
`pheatmap`, `matrixStats`, `microbenchmark`.

From Bioconductor: `DESeq2`, `EnhancedVolcano`.

```{r verify}
# Check what you ended up with at any point:
for (p in c("ggplot2", "ggpubr", "dplyr", "RColorBrewer", "readxl", "openxlsx",
            "pheatmap", "matrixStats", "microbenchmark",
            "DESeq2", "EnhancedVolcano")) {
  cat(sprintf("%-18s %s\n", p, tryCatch(as.character(packageVersion(p)),
                                        error = function(e) "not installed")))
}
```
