---
title: "Installation and setup — R track"
author: "Enrique Audain Martinez"
date: "`r Sys.Date()`"
output:
   html_document: default
   pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What this course is pinned to

| | |
|---|---|
| R | **4.6.x** (recommended) |
| Bioconductor | **3.23** |
| CRAN snapshot | **2026-09-01** |

Install R 4.6.x from [cran.r-project.org](https://cran.r-project.org) and RStudio
Desktop from [posit.co](https://posit.co/download/rstudio-desktop/).

### If you already have an older R

The script still works on R 4.1 through 4.5. It cannot give you Bioconductor 3.23
though — Bioconductor pairs each release with one R version, and that pairing is
not ours to override:

| Your R | You get Bioconductor | DESeq2 |
|---|---|---|
| 4.1 | 3.14 | 1.34 |
| 4.2 | 3.16 | 1.38 |
| 4.3 | 3.18 | 1.42 |
| 4.4 | 3.20 | 1.46 |
| 4.5 | 3.22 | 1.50 |
| **4.6** | **3.23** | **1.52** |

Everything installs and the lessons run, but results and plots may differ from the
committed outputs, and the further back you are the more they differ. The script
prints a warning telling you which pairing you got.

If you can install 4.6, do — R versions sit side by side quite happily, so nothing
you already have is disturbed, and you switch between them in RStudio under
**Tools → Global Options → General → R version**.

## Install everything

Open `ZebraQ.Rproj` in RStudio, then run this once:

```{r install, eval=FALSE}
# Works whether the working directory is the project root (which it is when you
# open ZebraQ.Rproj) or this setup/ folder (which it is if you Run Chunk here).
source(if (file.exists("setup/install_r.R")) "setup/install_r.R" else "install_r.R")
```

That is the whole setup. The script checks your R version, installs the nine CRAN
packages and two Bioconductor packages the lessons use, and prints the version of
each so you can see that it worked.

## Why a script instead of `install.packages()`

Two problems come up on people's own laptops, and the script exists to avoid both.

**Everyone ends up with different versions.** A plain `install.packages("ggplot2")`
gives you whatever CRAN published today. Over a semester the course drifts apart
from itself, and a plot that works on one laptop looks different on another. The
script installs from a *dated snapshot* of CRAN, so everybody gets the same
versions no matter when they run it.

**R offers to compile from source, and compiling fails.** On macOS and Windows the
default `pkgType` is `"both"`, which means R asks

> Do you want to install from sources the package which needs compilation?

whenever the source version is newer than the available binary. Saying yes needs a
compiler toolchain — Rtools on Windows, the Xcode command line tools on macOS —
which most people do not have, and the install dies partway through. The script
sets `pkgType = "binary"` so the question never comes up.

One limit worth knowing: the CRAN packages are frozen to a date, but the two
Bioconductor packages are pinned only to a *release* (3.23). Bioconductor issues
patch updates within a release, so two people installing months apart can differ
in DESeq2's patch version. There is no dated Bioconductor snapshot to point at.

The packages come from [Posit Package Manager](https://packagemanager.posit.co),
which serves prebuilt binaries for Windows, macOS (both Intel and Apple Silicon)
and Linux. Plain CRAN has no Linux binaries at all, so a Linux install without this
would compile every package from source.

## If something goes wrong

**"You are on R 4.4, so you will get Bioconductor 3.20"** — a warning, not an
error. The install continues and the lessons run; see the table above for what
differs. Install R 4.6 if you want to match the committed outputs exactly.

**"You are running R 4.0, which this course cannot install for"** — this one does
stop. R 4.0 and older are outside the supported range; install R 4.6.x and select
it in RStudio as described above.

**"ggpubr, readxl, openxlsx not available for R 4.1 on this platform"** — older R
versions no longer have a complete set of prebuilt macOS binaries, because CRAN
stops building for them as new releases arrive. Those three packages have to be
compiled, which needs the Xcode command line tools. R 4.6 has the full set.

**A package still fails.** Run the script again; it only installs what is missing,
so it is safe to repeat. If the same package fails twice, send the last 20 lines of
output rather than a screenshot of the red text.

**You want to know exactly what you have.** `sessionInfo()` prints your R version
and every loaded package — the R equivalent of the `session_info.show()` you will
see at the end of each Python notebook.

## Packages this installs

From CRAN: `tidyverse` (required by S6), `ggplot2`, `ggpubr`, `dplyr`,
`RColorBrewer`, `readxl`, `openxlsx`, `pheatmap`, `matrixStats`, `microbenchmark`.

From Bioconductor: `DESeq2`, `EnhancedVolcano`.

```{r verify}
# Check what you ended up with at any point:
for (p in c("tidyverse", "ggplot2", "ggpubr", "dplyr", "RColorBrewer", "readxl",
            "openxlsx", "pheatmap", "matrixStats", "microbenchmark",
            "DESeq2", "EnhancedVolcano")) {
  cat(sprintf("%-18s %s\n", p, tryCatch(as.character(packageVersion(p)),
                                        error = function(e) "not installed")))
}
```
